---
title: "Linear functionals; or: what to do when you've got a bunch of x-values and only one y"
author: Eric Pedersen
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
    highlight: haddock
---

##1. Background

Up to now, we've been looking at data where every coefficient we're using to 
predict our y-values consists of a single value for each y. However, not all 
predictor data looks like this. In many cases, a single coefficient may have 
several values associated with the same outcome variable. For instance, Let's
say we've measured total ecosystem production in several different lakes, as
well as temperature along a gradient away from shoreline in the lake. We want to know:

1. how well temperature predicts production
2. whether temperatures at different distances from the shore have different effects on
production (as production varies a lot between the littoral and pelagic zone).

Now, there looks like there might be a few ways to answer this in a standard gam
setting. First we could just average temperatures across locations and use that 
as a predictor to answer question 1. However, that doesn't give us any insight 
into question 2, and it's pretty easy to imagine a case like warmer temperatures
at the shore increase production, but warmer temperatures in the middle of the 
lake have very little effect, or where a single warm area could substantially
increase production even in a cold lake. We could also try to fit a seperate
smooth of production on temperature at each shore distance. However, if we have
a lot of distances, this ends up fitting a bunch of seperate smooths, and we'd
rapidly run out of degrees of freedom. Also, it would mean throwing away
information; we'd expect that the effect of temperature a meter below from the
shore should be very similar to the effect of temperature right at the
shoreline, but by fitting a seperate smooth for each, we're ignoring that (and
likely going to suffer from concurvity issues to boot!).

What we need is a method that can account for the fact that we have multiple x 
values for a given predictor. Fortunately, `mcgv` can handle this pretty easily.
It does this by allowing to pass matrix-valued predictors to `s()` and `te()` 
terms. When you pass `mgcv` a matrix, it will fit the same smooth function to 
each of the columns of the matrix, then sum across the rows of the matrix of
transformed values to give the estimated mean value for a given `y`. Let's say
`y` is a one-dimensional outcome, with n measured values, and `x` is a matrix
with n rows and k columns. The predicted values for the ith value of `y` would
be: $y_i \sim\sum_{j=1}^k f(x_{i,j})$, where $f(x_{i,j})$ is itself a sum of
basis functions multiplied by model coefficients as before.

In mathemetical terms, this makes our smooth term a *functional*, where our
outcome is a function of a vector of terms rather than a single term, and a
linear functional, as it's a linear sum of smooth terms for each column; but you
really don't need to know much about functionals to use these. I only mention
the functional thing because if you want to dig into this approach further or if
you want to read help files on these, as you have to search
`?linear.functional.terms`.

There's a few major useages Ive found for these. There's likely more than this,
but these are all cases I've encountered. Also, each of these cases requires you
to set up your predictor matrices carefully, so make sure you know what kind of
functional you'll be fitting, and what predictor matrix is going where in the `s()`
function.

### Case 1. Nonlinear averaging. 
The first case, and the simplest to fit, is to estimate 
a nonlinear average. Let's look at the lake example from before. Say we think 
that distance from shore shouldn't matter, but that production increases
non-linearly with temperature. In that case, a cold lake with a few hot spots
may actually be more productive than a lake that's consitently luke warm.
Therefore, if you average over temperatures before estimating a function, you'll 
miss this effect (as we know from statistics that the average of a non-linear
function applied to x does not, in general, equal the function applied to the
average of x). In this case, we can just provide a matrix of values (the predictor matrix) to `s()`: in this case, the predictor matrix is a matrix of temperatures where each column is the temperature measured at one of our sites. 
We do this in the code below with the variable `temp_matrix`. 


```{r, echo=T,tidy=F,include=T, message=FALSE,highlight=TRUE}
library(dplyr)
library(mgcv)
library(ggplot2)
library(tidyr)
n_lakes = 200
lake_mean_temps = rnorm(n_lakes, 25,6)
lake_data = as.data.frame(expand.grid(lake = 1:n_lakes,
                                      shore_dist_m = c(0,1,5,10,20,40)))
lake_data$temp = rnorm(n_lakes*6, lake_mean_temps[lake_data$lake], 8)
lake_data$production = 5*dnorm(lake_data$temp, 30, sd = 5)/dnorm(1,0,5)+rnorm(n_lakes*6, 0, 1)
lake_data = lake_data %>% group_by(lake)%>%
  mutate(production = mean(production),
         mean_temp = mean(temp))%>%
  spread(shore_dist_m, temp)


lake_data$temp_matrix = lake_data %>% 
  select(`0`:`40`) %>%
  as.matrix(.)
head(lake_data)


mean_temp_model = gam(production~s(mean_temp),data=lake_data)
nonlin_temp_model = gam(production~s(temp_matrix),
                                data=lake_data)

layout(matrix(1:2, nrow=1))
plot(mean_temp_model)
plot(nonlin_temp_model)
layout(1)
temp_predict_data = data.frame(temp = seq(10,40,length=100),
                          mean_temp = seq(10,40,length=100))

temp_matrix = matrix(seq(10,40,length=100),nrow= 100,ncol=6,
                                       byrow = F) 

temp_predict_data = temp_predict_data %>%
  mutate(linear_average = as.numeric(predict(mean_temp_model,.)),
         nonlinear_average = as.numeric(predict(nonlin_temp_model, .)),
        true_function = 5*dnorm(temp, 30, sd = 5)/dnorm(1,0,5))%>%
  gather(model, value,linear_average:true_function)

#This plots the two fitted models vs. the true model.
ggplot(aes(temp, value,color=model), data=temp_predict_data)+
  geom_line()+
  scale_color_brewer(palette="Set1")+
  theme_bw(20)
```

The nonlinear model is substantially more uncertain at the ends,
bu it fits the true function much more effectively. The function from 
the average data substantially underestimates the effect of temperature on 
production.

### Case 2: Weighted averaging
The next major case where linear functionals are useful is for weighted 
averages. Here, you have some predictor variable measured at various points at
different distances or lags from a given point, or at different locations along
some gradient. This could be a variable that's been meaured at various distances
away from each observed site, and you want to understand at what scale that
variable will affect your parameter of interest. It could also be a predictor
variable measured at several time points before the observation, and you want to
know what at what lags the two variables interact.

In this case, we'll assume that the relationship between our variable of 
interest (x) and the outcome is linear at any given lag, but that linear
relationship changes smoothly with the lag. We'll look at the case where both
relationships are nonlinear next. Here we have to create two matrices. The first
is the lag matrix, and it will have one column for each lag we're interested in. 
All the values in a given column will be equal to the lag value.
The second matrix is the predictor matrix, and it is the same as (see [the section below](#by_var)) on `by=` terms on how this works).


```{r, echo=T,tidy=F,include=T, message=FALSE,highlight=TRUE}
library(dplyr)
library(mgcv)
library(ggplot2)
library(tidyr)
n_lakes = 200
lake_mean_temps = rnorm(n_lakes, 25,6)
lake_data = as.data.frame(expand.grid(lake = 1:n_lakes,
                                      shore_dist_m = c(0,1,5,10,20,40)))
lake_data$temp = rnorm(n_lakes*6, lake_mean_temps[lake_data$lake], 8)
lake_data$production = with(lake_data, rnorm(n_lakes*6, 
                             temp*3*exp(-shore_dist_m/10),1))
lake_data = lake_data %>% group_by(lake)%>%
  mutate(production = mean(production),
         mean_temp = mean(temp))%>%
  spread(shore_dist_m, temp)

#This is our lag matrix for this example
shore_dist_matrix = matrix(c(0,1,5,10,20,40), nrow=n_lakes,ncol=6, byrow = T)
head(shore_dist_matrix)

#This is our predictor matrix
lake_data$temp_matrix = lake_data %>% 
  select(`0`:`40`) %>%
  as.matrix(.)
head(lake_data)

#Note: we need to set k=6 here, as we really only have 6 degrees of freedom (one
#for each distance we've measured from the shore.

nonlin_temp_model = gam(production~s(shore_dist_matrix, by= temp_matrix,k=6),
                                data=lake_data)
plot(nonlin_temp_model)

#This plots the two fitted models vs. the true model.
ggplot(aes(temp, value,color=model), data=temp_predict_data)+
  geom_line()+
  scale_color_brewer(palette="Set1")+
  theme_bw(20)
```

## 2. Key concepts and functions

### using `by=` terms in smoothers {#by_var}
One of the most useful features in `mgcv` is the `by=` argument for smooths. 
This has two uses. 

1. For a given smooth (say, `y~s(x)`), if you set `s(x,by=group`), where group
is some factor-leveled predictor, `mgcv` will fit a seperate smooth of x for
each level of that factor. The model will produce a different smooth $s_k(x)$
for each kth level, allowing you to test if a given relationship varies betewen
group. 
2. If you instead you set `s(x, by=z)` where z is a numerical value, `mgcv` will
fit a varying slopes regression. Instead of modeling y as a smooth function of
x, this will model y as a *linear* function of z, where the slope of the
relationship between z and y changes smoothly as a function of x, so the 
predicted value for the ith value of y would be:  $y_i \sim  f(x_i) \cdot z_i$. 
If you're using matrix predictors as discussed above, for numerical predictors 
the `by` variable `z` also has to be a matrix with the same dimensions as `x`. 
The predicted value in this case will be: $y_i \sim  \sum_{j=1}^k f(x_{i,j})\cdot z_{i,j}$, where $j$ are the columns of the matrices $x$ and $z$.


##Calculating dispersal kernels

```{r, echo=T,tidy=F,include=T, message=FALSE,highlight=TRUE}
yukon_seedling_data = read.csv("data/yukon_seeds/seed_data.csv")
yukon_source_data  =read.csv("data/yukon_seeds/seed_source_locations.csv")
seed_dist = matrix(0, nrow = nrow(yukon_seedling_data), 
                   ncol= nrow(yukon_source_data))
for(i in 1:nrow(yukon_seedling_data)){
    seed_dist[i,] = sqrt((yukon_source_data$X- yukon_seedling_data$X[i])^2 + (yukon_source_data$Y- yukon_seedling_data$Y[i])^2)
}
seed_dist_l = log(seed_dist)
yukon_seedling_data$min_dist_l = apply(seed_dist_l, MARGIN = 1,min)

basic_dispersal_model = gam(n_spruce~s(min_dist_l)+offset(log(plot_area_m2)),
                            data=yukon_seedling_data, family=nb)
full_dispersal_model = gam(n_spruce~s(seed_dist_l)+offset(log(plot_area_m2)),
                            data=yukon_seedling_data, family=nb)
```

