---
title: "Spatio-temporal dynamics is for the birds"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
    highlight: haddock
---

##1. Background 
We've talked about modelling spatial trends in the main workshop and in an
extended example, and about the trials and tribulations of time series data
in another extended example. Here, we're going to combine these, to look at 
one of the most challenging types of data for linear models to deal with:
spatiotemporal data. A spatiotemporal dataset is one where the outcome of
interest has been observed at multiple locations and at multiple points in time,
and we want to know how temporal trends change across the landscape (or, from
the other view, how spatial patterns are changing over time). 

We don't have time to really get into the nuts and bolts issues around fitting 
complex spatio-temporal models. This exercise is instead designed to give you
a feel for how to use `mgcv` to model complex interactions between variables
with different natural scales, and how to include simple random effects into 
gam models.

## 2. Key concepts and functions: 
Here's a few key ideas and R functions you should familiarize yourself with if 
you haven't already encountered them before. For the R functions (which will be)
highlighted `like this`, use `?function_name` to look them up. 

### Tensor product (`te`) smooths
This tutorial will rely on tensor product smooths. These are used to model 
nonlinear interactions between variables that have different natural scales. The
interaction smooths we saw in the start of the workshop were thin plate splines 
`y~s(x1,x2, bs="tp")`). These splines assume that the relationship between y and
x1 should be roughly as "wiggly" as the relationship between y and x2, and only
has one smoothing parameter. This makes sense if x1 and x2 are similar variables
(say latitude and longitude, or arm length and leg length), but if we're looking
at the interacting effects of, say, temperature and body weight, a single
smoothing term doesn't make sense.

This is where tensor product smooths come in. These assume that each variable of
interest should have its own smooth term. These create multi-dimensional smooth 
terms by multiplying the values of each one-dimensional basis by each value of
the second. This is a pretty complicated idea, and hard to describe effectively 
without more math than we want to go into here. But you can get the idea from a
simple example. Let's say we have two basis functions, one for variable x1 and 
one for variable x2. The function for x1 is a bell-curve shaped, and the one for
x2 is linear. The basis functions will look like this:

```{r, echo=T,tidy=F,results="hide", include=T, message=FALSE,highlight=TRUE}
library(mgcv)
library(ggplot2)
library(dplyr)
library(viridis)
set.seed(1)
n=25
dat = as.data.frame(expand.grid(x1 = seq(-1,1,length.out = n),
                                x2= seq(-1,1,length=n)))
dat$x1_margin = -dat$x1
dat$x2_margin = dnorm(dat$x2,0,0.5)
dat$x2_margin = dat$x2_margin - mean(dat$x2_margin)

layout(matrix(1:2,ncol=2))
plot(x1_margin~x1,type="l", data= dat%>%filter(!duplicated(x1)))
plot(x2_margin~x2,type="l", data= dat%>%filter(!duplicated(x2)))
layout(1)
```

However, the tensor product of the two variables is more complicated. Note that
when x1 is below zero, the effect of x2 is positive at middle values and negative
at low values, and vice versa when x1 is above zero. 
```{r, echo=T,tidy=F,results="hide", include=T, message=FALSE,highlight=TRUE}
dat$x1x2_tensor = dat$x1_margin*dat$x2_margin
ggplot(aes(x=x1,y=x2,fill=x1x2_tensor),data=dat)+
  geom_tile()+
  scale_x_continuous(expand=c(0,0))+
  scale_y_continuous(expand=c(0,0))+
  scale_fill_viridis()+
  theme_bw()
```

A full tensor product smooth will consist of several basis functions constructed
like this. The resulting smooth is then fitted using the standard gam model, but
with multiple penalty terms (one for each term in the tensor product). Each
penalty penalizes how much the function deviates from a smooth function in the
direction of one of the variables, averaged over all the other variables. This
allows us to have the function be more or less wiggly in one direction than
another. 

They are built using the `te()` function, like this:

```{r, echo=T,tidy=F, include=T, message=FALSE,highlight=TRUE}
# We'll recycle our fake marginal and tensor products as the expected value for
# a simulation. Waste not, want not!
dat$y = rnorm(n^2, dat$x1x2_tensor+dat$x1_margin+dat$x2_margin,0.5) 

te_demo = gam(y~te(x1, x2, k = 4^2),data=dat,method="REML") #k=5^2 so there will be 5 marginal bases for each dimension
plot.gam(te_demo,scheme = 2)
summary(te_demo)
print(te_demo$sp) 
#These are the estimated smooths for the two dimensions. Note that the smooth for
#x1 (the linear basis) is way smoother than the one for x2
```



### Tensor interaction smooths
There are many cases when we want an estimate of both the average effect of each
variable as well as the interaction. This is tough to extract from just using 
`te()` terms, but if we try to combine the two (say, with 
`y~s(x1)+s(x2)+te(x1,x2))`) the result is often numerically unstable; it won't
give reliable consistent separation into the terms we want. Instead, we use `ti`
terms for the interaction. The `ti` terms are set up so the average effect of
each variable in the smooth is already excluded from the interaction (and they'll
have fewer basis functions than the equivalent `te` term). Search `?te` if you're 
curious about how `mcgv` sets these up; for now, we'll just use them as a black 
box. 

This is how you use them in practice: 

```{r, echo=T,tidy=F, include=T, message=FALSE,highlight=TRUE}
# We'll recycle our fake tensor product as the expected value for a simulation
# Waste not, want not!

ti_demo = gam(y~s(x1,k=4)+s(x2,k=4) +ti(x1, x2, k = 4^2),
              data=dat,method="REML")
layout(matrix(1:3,nrow=1))
plot.gam(ti_demo,scheme = 2)
layout(1)
summary(ti_demo)
print(ti_demo$sp) 
#These are the estimated smooths for the two dimensions. Note that the smooth for
#x1 (the linear basis) is way smoother than the one for x2
```

Note that it does a pretty good job of extracting our assumed marginal 
basis functions (linear and Gaussian) and the tensor product into three clearly 
separate functions. 

NOTE: Never use `ti` on its own without including the marginal smooths! This
is for the same reason you shouldn't exclude individual linear terms but include
interactions in `glm` models. It really virtually never makes sense to talk about a function with a strong interaction but not marginal effects.

### Random effects smooths
The last smooth type we'll use in this exercise is the random effects smooth. 
This is just what it sounds on the box: it's a smooth for factor terms, with one
coefficient for each level of the group (the basis functions for this case), and 
a penalty term on the squared value of the coefficients to draw them towards the
global mean. This is exactly the same mechanism that packages like `lme4` and
`nlme` use to calculate random effects. In fact, there's a very deep connection
between gams and mixed effects,but we won't go into that here. 

```{r, echo=T,tidy=F, include=T, message=FALSE,highlight=TRUE}
n= 10
n_groups= 10
group_means = rnorm(n_groups, 0, 2)
dat = data_frame(group = rep(1:n_groups, each=n),
                 group_f = factor(group))
dat$y = rnorm(n*n_groups, 3+group_means[dat$group],2)
re_demo = gam(y~s(group_f, bs="re"),
              data=dat,method="REML")
plot.gam(re_demo,scheme = 2)
summary(re_demo)
print(re_demo$sp) 
```

It's important to know when using `bs="re"` terms whether the variable you're 
giving it is numeric or factorial. `mgcv` will not raise an error if you give it
numerical variables to `bs="re"`. Instead, it'll just fit a linear model with a 
penalty on the slope of the line between x and y. Try it with the data about to 
see, by switching `s(group_f, bs="re")` with `s(group, bs="re")`. `bs="re"` is 
set up like this to allow you to use random effects smooths to fit things like 
varying slopes random effects models, but make sure you know what you're doing
when venturing into this.

## 3. Spatiotemporal models
Now that we've covered the key smoothers, let's take a look at how we'd use them
to model spatiotemporal data. For this section, we'll use data from the Florida 
routes of the Breeding Bird Survey. If you haven't heard of it before, this is a
long-term monitoring program tracking bird abundance and diversity across the US 
and Canada. It consists of many (over 4000) 24.5 mile long routes. Trained 
observers walk these routes each year during peak breeding season, and every 0.5
miles, they stop and count and identify all the birds they can see or hear in a
0.25 mile radius of the stop. This is one of the best long-term data sets we
have on bird diversity and abundance, and has been used in countless papers. 

Here, we're going to use this data to answer a couple relatively simple
questions: how does bird species richness vary across the state of Florida, has
average richness changed over time, and has the spatial pattern of richness
varied? We obtained data from [this
site](https://www.pwrc.usgs.gov/bbs/RawData/), and summarized it into the total
number of different species and the total number of individual birds observed on
each route in each year in the state of Florida. 

First, we'll load the data, and quickly plot it:
```{r, echo=T,tidy=F, include=T, message=FALSE,highlight=TRUE}
library(maps)
florida_map = map_data("state","florida")%>%
  transmute(Longitude= long, Latitude = lat, order=order)
florida_birds = read.csv("data/bbs_data/bbs_florida_richness.csv")
head(florida_birds)

ggplot(aes(x=Year,y=Richness),data=florida_birds)+geom_point()+
  theme_bw()
ggplot(aes(x=Longitude,y=Latitude,col=Richness), data=florida_birds) +
    geom_polygon(data=florida_map,fill=NA, col="black")+
  geom_point()+theme_bw()
```

It looks like richness may have declined in the end of the time series, and
there seems to be somewhat of a north-south richness gradient, but we need to
model it to really tease these issues out. 

### Modelling space and time seperately
We'll first take a stab at this to determine if these mean trends actually come
out of the model. As richness is a count variable
we'll use a Poisson gam to model it. This means
that we're assuming mean richness is the product of a smooth
function of location and of date, with no interactions; 
E(richness) = exp(f(year))*exp(g(location)).

```{r, echo=T,tidy=F, include=T, message=FALSE,highlight=TRUE}
richness_base_model = gam(Richness~ s(Longitude,Latitude)+s(Year),
                          data=florida_birds,family=poisson,method="REML")
layout(matrix(1:2,ncol=2))
plot(richness_base_model,scheme=2)
layout(1)
summary(richness_base_model)
```

Note that we do have a thin-plate interaction between longitude and latitude.
This is assuming that the gradient should be equally smooth latitudinally and
longitudinally. This may not be totally accurate, as the function seems to
change faster latitudinally, but we'll ignore that until the exercises at the 
bottom. The model seems to be doing pretty well; it captured the original patterns
we saw, and explains a not-inconsiderable 45% of the deviance. 

Let's see if there's any remaining spatiotemporal patterns though:

```{r, echo=T,tidy=F, include=T, message=FALSE,highlight=TRUE}
source("code_snippets/quantile_resid.R")

florida_birds$base_resid = rqresiduals(richness_base_model)

ggplot(aes(Longitude, Latitude),
       data= florida_birds %>% filter(Year%%6==0))+ #only look at every 6th year
  geom_point(aes(color=base_resid))+
  geom_polygon(data=florida_map,fill=NA, col="black")+
  scale_color_viridis()+
  facet_wrap(~Year)+
  theme_bw()

```
There's definitely still a pattern in the residuals, such as higher than
expected diversity in the mid-latitudes during the 1990's.

### Joint models of space and time 
The next step is to determine if there is a significant interaction. This is
where tensor product splines come in. Here, we'll use the `ti` formulation, as
it allows us to directly test whether the interaction significantly improves our
model.

Something important to note: the `d=c(2,1)` argument it `ti`. This tells the
function that the smooth should consist of tensor product between a
2-dimensional smooth (lat-long) and a 1-dimensional term (Year).
 
```{r, echo=T,tidy=F, include=T, message=FALSE,highlight=TRUE}
richness_ti_model = gam(Richness~ s(Longitude,Latitude)+s(Year) + 
                          ti(Longitude, Latitude, Year, d=c(2,1)),
                          data=florida_birds,family=poisson,method="REML")
layout(matrix(1:2,ncol=2))
plot(richness_ti_model,scheme=2)
layout(1)
summary(richness_ti_model)

```
This improves the model fit substantially, which you can see by directly
comparing models with `anova`:
```{r, echo=T,tidy=F, include=T, message=FALSE,highlight=TRUE}
anova(richness_base_model,richness_ti_model,test = "Chisq")

```


The question then becomes: what does this actually mean? Where are we seeing the
fastest and slowest rates of change in richness? It's often difficult to
effectively show spatiotemporal changes, partially as it would require a
4-dimensional graph to really show what's going on. Still, we can look at a
couple different views by using `predict` effectively:
```{r, echo=T,tidy=F, include=T, message=FALSE,highlight=TRUE}
#First we'll create gridded data
predict_richess = expand.grid(
  Latitude= seq(min(florida_birds$Latitude), 
                max(florida_birds$Latitude),
                length=50),
  Longitude = seq(min(florida_birds$Longitude),
                  max(florida_birds$Longitude),
                  length=50),
  Year = seq(1970,2015,by=5)
)
# This now selects only that data that falls within Florida's border
predict_richess = predict_richess[with(predict_richess, 
                                       inSide(florida_map, Latitude,Longitude)),]
predict_richess$model_fit = predict(richness_ti_model,
                                    predict_richess,type = "response")
ggplot(aes(Longitude, Latitude, fill= model_fit),
       data=predict_richess)+
  geom_tile()+
  facet_wrap(~Year,nrow=2)+
  scale_fill_viridis("# of species")+
  theme_bw(10)
```


Another way of looking at this is by estimating the rate of change
at each location at each time point, and figuring out which locations
are changing most quickly:

```{r, echo=T,tidy=F, include=T, message=FALSE,highlight=TRUE}

predict_richess$model_change =predict(richness_ti_model,
                                      predict_richess%>%mutate(Year=Year+1),
                                      type = "response") -
  predict_richess$model_fit 
ggplot(aes(Longitude, Latitude, fill= model_change),
       data=predict_richess)+
  geom_tile()+
  facet_wrap(~Year,nrow=2)+
  scale_fill_gradient2("Rate of change\n(species per year)")+
  theme_bw(10)
```

### Accounting for local heterogeneity with `bs="re"`
Examining this, it seems to indicate that richness fluctuated across the state
throughout the study period, but then began sharply declining state-wide in the
2000's. (Disclaimer: I don't know enough about either the BBS or Florida to say
how much of this decline is real! Please don't quote me in newspapers saying
biodiversity is collapsing in Florida).

There is still one persistent issue: viewing the maps of richness, there appear
to be small hotspots and coldspots of richness across the state. This may be due
to a factor I've ignored so far: the data is collected in specific, repeated
routes. If it's just easier to spot birds on one route, or there's a lot of
local nesting habitat there, it will have higher richness than expected over the
whole study. Adjusting for these kinds of hot and cold spots will force our
model to be overly wiggly. Further, not all routes are sampled every year, so we
want to make sure that our trends don't just result from less diverse routes
getting sampled later in the survey.

To try and account for this effect (and other grouped local factors), we 
can use random effects smooths. That would make our model look like this:

```{r, echo=T,tidy=F, include=T, message=FALSE,highlight=TRUE}
florida_birds$Route_f = factor(florida_birds$Route)
richness_ti_re_model = gam(Richness~ s(Longitude,Latitude)+s(Year) + 
                          ti(Longitude, Latitude, Year, d=c(2,1))+
                          s(Route_f, bs="re"),
                          data=florida_birds,family=poisson,method="REML")
layout(matrix(1:3,nrow=1))
plot(richness_ti_re_model,scheme=2)
layout(1)

summary(richness_ti_re_model)
anova(richness_ti_model,richness_ti_re_model)
```

We can view the same plots as before with our new model:
```{r, echo=T,tidy=F, include=T, message=FALSE,highlight=TRUE}
predict_richess$Route_f = 1
predict_richess$model_fit = predict(richness_ti_re_model,
                                    predict_richess,type = "response")
predict_richess$model_change =predict(richness_ti_re_model,
                                      predict_richess%>%mutate(Year=Year+1),
                                      type = "response") -
  predict_richess$model_fit 

ggplot(aes(Longitude, Latitude, fill= model_fit),
       data=predict_richess)+
  geom_tile()+
  facet_wrap(~Year,nrow=2)+
  scale_fill_viridis("# of species")+
  theme_bw(10)

ggplot(aes(Longitude, Latitude, fill= model_change),
       data=predict_richess)+
  geom_tile()+
  facet_wrap(~Year,nrow=2)+
  scale_fill_gradient2("Rate of change\n(species per year)")+
  theme_bw(10)
```
This definely seems to have levelled off some of the hot spots from before, 
implying that bird richness is more evenly distributed than our prior model 
suggested. Still, the large-scale trends remain the same from before.

## 4. Exercises

1. Given that most of the variability appears to be latitudinal, try to adapt 
the tensor smoothers so that latitude,longitude, and year all have their own 
penalty terms. How does this change model fit? 

2. Following up on the prior
exorcise: build a model that allows you to explicitly calculate the marginal
latitudinal gradient in diversity, and to determine how the latitudinal gradient
has changed over time. 

3. We know from basic sampling theory that if you sample
fewer individuals over time, you also expect to find lower richness, even if
actual richness stays unchanged. What does the spatiotemporal pattern of
abudance look like? Does adding bird abudance as a predictor explain any of the
patterns in richness?
